---
title: "Homework 7"
author: "Gwen Ellis"
date: "2024-03-01"
output: html_document
---

```{r, include=FALSE}
# packages
library(ggplot2)
library(MASS)
library(tidyverse)
library(gridExtra)
```


# Simulating allele frequencies with a beta simulation

Scenario: In a previous GWAS, several loci were implicated in thermal tolerance resulting in different allele frequencies for the SNPs in different populations with different environments through local adaptation. These same loci are now being assessed in three new populations to see if the pattern of local adaptation holds true. The expectation is that the population in the warmest environment ("hot") will have an average higher allele frequency across all loci in comparison to the "warm" and "cold" environment populations.
Goal: ANOVA for average allele frequencies across the loci in 3 populations

For this homework, I am simulating allele frequencies for an adaptive trait at n number of loci using a beta distribution. In this case, I will be tracking the minor allele frequency where the minor allele is the adaptive one. This means that we expect the average allele frequency to be relatively low (~5-10%). This number will be bounded by zero and 0.05 as it is still being considered the minor allele. In this simulation, shape1 is similar to how many times we see the minor allele and shape2 is similar to how many times we see the major allele. For a beta distribution, the mean = shape1/(shape1 + shape2) and the variance is proportional to the magnitude of the two parameters. A larger magnitude results in less variance and smaller magnitudes result in more variance. For this simulation, there should be a very small mean (0.05 - 0.15) and relatively low variance as most of the loci should have similar allele frequencies.

## TRIAL 1 - First set of allele frequency distributions with random numbers from beta distribution 
```{r, message=FALSE, warning=FALSE}
# beta distribution (has to be bounded by 0 and 1 because they're allele frequencies), number of loci = 100, mean will be based on thermal tolerance

popname <- c("cold","warm","hot") # population names
n_loci <- 100 # number of loci, how polygenic is the trait
shape1_pops <- c(2,3,5) # shape1 for beta distribution
shape2_pops <- c(50,50,50) # shape2 for beta distribution
names(shape1_pops) <- popname
names(shape2_pops) <- popname

set.seed(3)
# allele frequency distributions for each population
cold_freq <- rbeta(n=n_loci, shape1=shape1_pops["cold"], shape2=shape2_pops["cold"]) 
warm_freq <- rbeta(n=n_loci, shape1=shape1_pops["warm"], shape2=shape2_pops["warm"])
hot_freq <- rbeta(n=n_loci, shape1=shape1_pops["hot"], shape2=shape2_pops["hot"])

# plot the allele frequencies for each population
cold_plot <- qplot(x=cold_freq, xlim=c(0,1), ylim=c(0,50),
                   color=I("black"), fill=I("cornflowerblue"),
                   main="Cold population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
warm_plot <- qplot(x=warm_freq, xlim=c(0,1),ylim=c(0,50),
                   color=I("black"), fill=I("goldenrod1"),
                   main="Warm population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
hot_plot <- qplot(x=hot_freq, xlim=c(0,1), ylim=c(0,50),
                  color=I("black"), fill=I("tomato"),
                  main="Hot population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
grid.arrange(cold_plot, warm_plot, hot_plot)

allele_freq <- c(cold_freq, warm_freq, hot_freq) # compile the allele frequencies
id <- 1:(sum(n_loci))
temp <- rep(popname, each=n_loci) # to make a column with all the populations
aov_df <- data.frame(id, temp, allele_freq) # put the id, pop temp, and allele frequencies into one df

# what are the average allele frequencies for each population?
avg_freq <- list(mean(cold_freq), mean(warm_freq), mean(hot_freq))
names(avg_freq) <- popname
print(avg_freq) 

# run the ANOVA
aov_mod <- aov(allele_freq~temp, data=aov_df) 
print(summary(aov_mod))
aov_list <- unlist(summary(aov_mod))
aov_summary <- list(F_value=aov_list[7], P_value=aov_list[9])


# making a graph, are there significant differences in allele frequency between the populations?
aov_df$temp <- factor(aov_df$temp , levels=c("cold", "warm", "hot"))
anova_plot1 <- ggplot(data=aov_df,aes(x=temp,y=allele_freq,fill=temp)) +
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "tomato")) +
  geom_boxplot()
print(anova_plot1)

#Is the difference in allele frequency between the populations significant?
if (aov_list[9] < 0.05) {
  return("YES")
} else {
  return("NO")
}

```

## TRIAL 2 - Different set of random numbers from beta distribution
```{r, message=FALSE, warning=FALSE}

# Distribution parameters
n_loci <- 100 
shape1_pops <- c(2,3,5) 
shape2_pops <- c(50,50,50) 
names(shape1_pops) <- popname
names(shape2_pops) <- popname

set.seed(7)
# allele frequency distributions for each population
cold_freq <- rbeta(n=n_loci, shape1=shape1_pops["cold"], shape2=shape2_pops["cold"]) 
warm_freq <- rbeta(n=n_loci, shape1=shape1_pops["warm"], shape2=shape2_pops["warm"])
hot_freq <- rbeta(n=n_loci, shape1=shape1_pops["hot"], shape2=shape2_pops["hot"])

# plot the allele frequencies for each population
cold_plot <- qplot(x=cold_freq, xlim=c(0,1), ylim=c(0,50),
                   color=I("black"), fill=I("cornflowerblue"),
                   main="Cold population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
warm_plot <- qplot(x=warm_freq, xlim=c(0,1),ylim=c(0,50),
                   color=I("black"), fill=I("goldenrod1"),
                   main="Warm population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
hot_plot <- qplot(x=hot_freq, xlim=c(0,1), ylim=c(0,50),
                  color=I("black"), fill=I("tomato"),
                  main="Hot population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
grid.arrange(cold_plot, warm_plot, hot_plot)

# df set up
allele_freq <- c(cold_freq, warm_freq, hot_freq) 
id <- 1:(sum(n_loci))
temp <- rep(popname, each=n_loci) 
aov_df <- data.frame(id, temp, allele_freq) 

# what are the average allele frequencies for each population?
avg_freq <- list(mean(cold_freq), mean(warm_freq), mean(hot_freq))
names(avg_freq) <- popname
print(avg_freq) 

# ANOVA
aov_mod <- aov(allele_freq~temp, data=aov_df) 
print(summary(aov_mod))
aov_list <- unlist(summary(aov_mod))
aov_summary <- list(F_value=aov_list[7], P_value=aov_list[9])

# Boxplot
aov_df$temp <- factor(aov_df$temp , levels=c("cold", "warm", "hot"))
anova_plot1 <- ggplot(data=aov_df,aes(x=temp,y=allele_freq,fill=temp)) +
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "tomato")) +
  geom_boxplot()
print(anova_plot1)

#Is the difference in allele frequency between the populations significant?
if (aov_list[9] < 0.05) {
  return("YES")
} else {
  return("NO")
}

```

## Changing the mean - not significant
```{r, message=FALSE, warning=FALSE}

# Distribution parameters
n_loci <- 100 
shape1_pops <- c(2,2,2) # if all populations have the same "count" of the minor allele, the means will be almost identical
shape2_pops <- c(50,50,50) 
names(shape1_pops) <- popname
names(shape2_pops) <- popname

set.seed(8)
# allele frequency distributions for each population
cold_freq <- rbeta(n=n_loci, shape1=shape1_pops["cold"], shape2=shape2_pops["cold"]) 
warm_freq <- rbeta(n=n_loci, shape1=shape1_pops["warm"], shape2=shape2_pops["warm"])
hot_freq <- rbeta(n=n_loci, shape1=shape1_pops["hot"], shape2=shape2_pops["hot"])

# plot the allele frequencies for each population
cold_plot <- qplot(x=cold_freq, xlim=c(0,1), ylim=c(0,50),
                   color=I("black"), fill=I("cornflowerblue"),
                   main="Cold population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
warm_plot <- qplot(x=warm_freq, xlim=c(0,1),ylim=c(0,50),
                   color=I("black"), fill=I("goldenrod1"),
                   main="Warm population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
hot_plot <- qplot(x=hot_freq, xlim=c(0,1), ylim=c(0,50),
                  color=I("black"), fill=I("tomato"),
                  main="Hot population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
grid.arrange(cold_plot, warm_plot, hot_plot)

# df set up
allele_freq <- c(cold_freq, warm_freq, hot_freq) 
id <- 1:(sum(n_loci))
temp <- rep(popname, each=n_loci) 
aov_df <- data.frame(id, temp, allele_freq) 

# what are the average allele frequencies for each population?
avg_freq <- list(mean(cold_freq), mean(warm_freq), mean(hot_freq))
names(avg_freq) <- popname
print(avg_freq) 

# ANOVA
aov_mod <- aov(allele_freq~temp, data=aov_df) 
print(summary(aov_mod))
aov_list <- unlist(summary(aov_mod))
aov_summary <- list(F_value=aov_list[7], P_value=aov_list[9])

# Boxplot
aov_df$temp <- factor(aov_df$temp , levels=c("cold", "warm", "hot"))
anova_plot1 <- ggplot(data=aov_df,aes(x=temp,y=allele_freq,fill=temp)) +
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "tomato")) +
  geom_boxplot()
print(anova_plot1)

#Is the difference in allele frequency between the populations significant?
if (aov_list[9] < 0.05) {
  return("YES")
} else {
  return("NO")
}

```



## Changing the mean - just barely significant
```{r, message=FALSE, warning=FALSE}

# Distribution parameters
n_loci <- 100 
shape1_pops <- c(1.3,1.5,1.9) # At this low magnitude, the values have to be very close together
shape2_pops <- c(50,50,50) 
names(shape1_pops) <- popname
names(shape2_pops) <- popname

set.seed(9)
# allele frequency distributions for each population
cold_freq <- rbeta(n=n_loci, shape1=shape1_pops["cold"], shape2=shape2_pops["cold"]) 
warm_freq <- rbeta(n=n_loci, shape1=shape1_pops["warm"], shape2=shape2_pops["warm"])
hot_freq <- rbeta(n=n_loci, shape1=shape1_pops["hot"], shape2=shape2_pops["hot"])

# plot the allele frequencies for each population
cold_plot <- qplot(x=cold_freq, xlim=c(0,1), ylim=c(0,50),
                   color=I("black"), fill=I("cornflowerblue"),
                   main="Cold population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
warm_plot <- qplot(x=warm_freq, xlim=c(0,1),ylim=c(0,50),
                   color=I("black"), fill=I("goldenrod1"),
                   main="Warm population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
hot_plot <- qplot(x=hot_freq, xlim=c(0,1), ylim=c(0,50),
                  color=I("black"), fill=I("tomato"),
                  main="Hot population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
grid.arrange(cold_plot, warm_plot, hot_plot)

# df set up
allele_freq <- c(cold_freq, warm_freq, hot_freq) 
id <- 1:(sum(n_loci))
temp <- rep(popname, each=n_loci) 
aov_df <- data.frame(id, temp, allele_freq) 

# what are the average allele frequencies for each population?
avg_freq <- list(mean(cold_freq), mean(warm_freq), mean(hot_freq))
names(avg_freq) <- popname
print(avg_freq) 

# ANOVA
aov_mod <- aov(allele_freq~temp, data=aov_df) 
print(summary(aov_mod))
aov_list <- unlist(summary(aov_mod))
aov_summary <- list(F_value=aov_list[7], P_value=aov_list[9])

# Boxplot
aov_df$temp <- factor(aov_df$temp , levels=c("cold", "warm", "hot"))
anova_plot1 <- ggplot(data=aov_df,aes(x=temp,y=allele_freq,fill=temp)) +
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "tomato")) +
  geom_boxplot()
print(anova_plot1)

#Is the difference in allele frequency between the populations significant?
if (aov_list[9] < 0.05) {
  return("YES")
} else {
  return("NO")
}

```


# Changing the number of alleles (n)
```{r, message=FALSE, warning=FALSE}
set.seed(13)
# Distribution parameters
n_loci <- 5 # need very few loci to start seeing insignificant difference
shape1_pops <- c(2,3,5) # 
shape2_pops <- c(50,50,50) 
names(shape1_pops) <- popname
names(shape2_pops) <- popname

# allele frequency distributions for each population
cold_freq <- rbeta(n=n_loci, shape1=shape1_pops["cold"], shape2=shape2_pops["cold"]) 
warm_freq <- rbeta(n=n_loci, shape1=shape1_pops["warm"], shape2=shape2_pops["warm"])
hot_freq <- rbeta(n=n_loci, shape1=shape1_pops["hot"], shape2=shape2_pops["hot"])

# plot the allele frequencies for each population
cold_plot <- qplot(x=cold_freq, xlim=c(0,1), ylim=c(0,50),
                   color=I("black"), fill=I("cornflowerblue"),
                   main="Cold population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
warm_plot <- qplot(x=warm_freq, xlim=c(0,1),ylim=c(0,50),
                   color=I("black"), fill=I("goldenrod1"),
                   main="Warm population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
hot_plot <- qplot(x=hot_freq, xlim=c(0,1), ylim=c(0,50),
                  color=I("black"), fill=I("tomato"),
                  main="Hot population allele frequencies",
                   ylab="Number of loci",
                   xlab="Allele frequency")
grid.arrange(cold_plot, warm_plot, hot_plot)

# df set up
allele_freq <- c(cold_freq, warm_freq, hot_freq) 
id <- 1:(sum(n_loci))
temp <- rep(popname, each=n_loci) 
aov_df <- data.frame(id, temp, allele_freq) 

# what are the average allele frequencies for each population?
avg_freq <- list(mean(cold_freq), mean(warm_freq), mean(hot_freq))
names(avg_freq) <- popname
print(avg_freq) 

# ANOVA
aov_mod <- aov(allele_freq~temp, data=aov_df) 
print(summary(aov_mod))
aov_list <- unlist(summary(aov_mod))
aov_summary <- list(F_value=aov_list[7], P_value=aov_list[9])

# Boxplot
aov_df$temp <- factor(aov_df$temp , levels=c("cold", "warm", "hot"))
anova_plot1 <- ggplot(data=aov_df,aes(x=temp,y=allele_freq,fill=temp)) +
  scale_fill_manual(values=c("cornflowerblue", "goldenrod1", "tomato")) +
  geom_boxplot()
print(anova_plot1)

#Is the difference in allele frequency between the populations significant?
if (aov_list[9] < 0.05) {
  return("YES")
} else {
  return("NO")
}
```
