---
title: "Homework7"
author: "Gwen Ellis"
date: "2024-03-04"
output: html_document
---

### libraries
```{r}
library(ggplot2)
library(MASS)
library(tidyverse)
```



For this homework, I am simulating allele frequencies for an adaptive trait at a single locus for n number of populations to see if there is a continuous relationship between allele frequency and the mean temperature of the warmest month (WorldClim BIO10) to see if there is local adaptation at this locus. In this case, I will be tracking the minor allele frequency where the minor allele is the adaptive one. This means that we expect the average allele frequency to be relatively low (10-20%). This number will be bounded by zero and 0.05 as it is still being considered the minor allele. From this homework, I expect that there will be a linear relationship between temperature and allele frequency.

## Trial 1, default values
```{r, message=FALSE, warning=FALSE}
n=100 # number of populations

# simulate allele frequencies for an adaptive trait at a single locus across 100 populations
allele_freq <- sort(rbeta(n=n, shape1=5, shape2=50))
qplot(x=allele_freq, xlim=c(0,1), color=I("black"), fill=I("cornflowerblue")) # checking how the distribution looks
print(mean(allele_freq))
print(var(allele_freq))

# shape1 (alpha) is similar to minor allele occurrence 
# shape2 (beta) is similar to major allele occurrence
# beta probability distribution will give minor allele frequency for the populations)
# mean is alpha/(alpha+beta)
# variance is approximate to how wide the distribution is and magnitude of shape1 and shape2

# simulate temperatures for 100 populations (this won't change)
set.seed(10)
temperature <- sort(runif(n=n, min=15, max=35)) # some hot temperatures, some cold, but average is ~25C
# mean = shape * scale
# variance = shape * scale^2
qplot(x=temperature, color=I("black"), fill=I("cornflowerblue"))

id <- c(1:n) # column with individual IDs

dist_df <- data.frame(id, temperature, allele_freq) # putting together the dataframe

df_lm <- lm(allele_freq~temperature, data=dist_df) # linear model for the data
print(summary(df_lm))
lm_list <- unlist(summary(df_lm))
model_sig <- list(intercept=lm_list$coefficients1, slope=lm_list$coefficients2, intercept_p=lm_list$coefficients7, slope_p=lm_list$coefficients8, r_sq=lm_list$r.squared)
print(model_sig)

sim_plot <- ggplot(data=dist_df, aes(x=temperature,y=allele_freq)) +
  xlim(15,35) +
  ylim(0,0.5) +
  geom_point() + 
  stat_smooth(method=lm, level=0.90) # default level=0.95 
print(sim_plot)

```




## Trial 2, default values - How does the simulated data change?
```{r, message=FALSE, warning=FALSE}
n=100 

allele_freq <- sort(rbeta(n=n, shape1=5, shape2=50))
qplot(x=allele_freq, xlim=c(0,1), color=I("black"), fill=I("cornflowerblue")) 
print(mean(allele_freq))
print(var(allele_freq))

set.seed(10)
temperature <- sort(runif(n=n, min=15, max=35)) 
qplot(x=temperature, color=I("black"), fill=I("cornflowerblue"))

id <- c(1:n) 

dist_df <- data.frame(id, temperature, allele_freq)

df_lm <- lm(allele_freq~temperature, data=dist_df) 
print(summary(df_lm))
lm_list <- unlist(summary(df_lm))
model_sig <- list(intercept=lm_list$coefficients1, slope=lm_list$coefficients2, intercept_p=lm_list$coefficients7, slope_p=lm_list$coefficients8, r_sq=lm_list$r.squared)
print(model_sig)

sim_plot <- ggplot(data=dist_df, aes(x=temperature,y=allele_freq)) +
  xlim(15,35) +
  ylim(0,0.5) +
  geom_point() + 
  stat_smooth(method=lm, level=0.90) 
print(sim_plot)

```


## Adjusting the mean 
The best way to adjust the mean for a beta distribution in this case is to increase/decrease shape1
```{r}
allele_freq <- sort(rbeta(n=n, shape1=20, shape2=50)) # changing from 5 to 20
qplot(x=allele_freq, xlim=c(0,1), color=I("black"), fill=I("cornflowerblue")) 
print(mean(allele_freq)) # The mean allele frequency is now much higher
print(var(allele_freq))

set.seed(10)
temperature <- sort(runif(n=n, min=15, max=35)) 
qplot(x=temperature, color=I("black"), fill=I("cornflowerblue"))

id <- c(1:n) 

dist_df <- data.frame(id, temperature, allele_freq)

df_lm <- lm(allele_freq~temperature, data=dist_df) 
print(summary(df_lm))
lm_list <- unlist(summary(df_lm))
model_sig <- list(intercept=lm_list$coefficients1, slope=lm_list$coefficients2, intercept_p=lm_list$coefficients7, slope_p=lm_list$coefficients8, r_sq=lm_list$r.squared)
print(model_sig) 

sim_plot <- ggplot(data=dist_df, aes(x=temperature,y=allele_freq)) +
  xlim(15,35) +
  ylim(0,0.5) +
  geom_point() + 
  stat_smooth(method=lm, level=0.90) 
print(sim_plot)

```

## Adjusting the variance
When just adjusting the variance, the relationship between temperature and allele frequency is maintained but just shifted. To break the relationship between these two variables, the variance in allele frequencies has to be increased. The best way to do this for a beta distribution is to dramatically decrease the magnitude of both shape parameters.
```{r}
allele_freq <- sort(rbeta(n=n, shape1=5, shape2=10)) # decreased from original by 1 order of magnitude
qplot(x=allele_freq, xlim=c(0,1), color=I("black"), fill=I("cornflowerblue")) 
print(mean(allele_freq)) # The mean allele frequency is now much higher
print(var(allele_freq))

set.seed(10)
temperature <- sort(runif(n=n, min=15, max=35)) 
qplot(x=temperature, color=I("black"), fill=I("cornflowerblue"))

id <- c(1:n) 

dist_df <- data.frame(id, temperature, allele_freq)

df_lm <- lm(allele_freq~temperature, data=dist_df) 
print(summary(df_lm))
lm_list <- unlist(summary(df_lm))
model_sig <- list(intercept=lm_list$coefficients1, slope=lm_list$coefficients2, intercept_p=lm_list$coefficients7, slope_p=lm_list$coefficients8, r_sq=lm_list$r.squared)
print(model_sig) 

sim_plot <- ggplot(data=dist_df, aes(x=temperature,y=allele_freq)) +
  xlim(15,35) +
  ylim(0,0.5) +
  geom_point() + 
  stat_smooth(method=lm, level=0.90) 
print(sim_plot)
```








