---
title: "class notes"
author: "Gwen"
date: "2024-01-25"
output: 
  html_document: 
    highlight: textmate
---

```{r}
# first chunk of code in rmd file
print(pi)
pi + 3

```


* `.Rmd` is Rstudio's special blend of Markdown (a text rendering language)
* `.md` is the more generic Markdown file type. In Rstudio, the intermediae .md fileas are not (in the default state) preserved

## Languages
* `LaTex` Typesetting language invented in the 1970s by computer programmer Donald Knuth for proper rendering of text and equations. Hundres of commands. Precise, but somewhat complicated to use.
* `Markdown` Greatly simplified version of LaTex with a small number of commands to master. Easy to use, texts look great.
* `yaml`^2^ Header material for `.Rmd` files that can be changed manually or by selecting options
* `R` Computer language for manipulating numbers, text, graphics, statistics

## Basic text formatting
* headers `#` to `######` *numbering from options
* text
* markup
  * *italic* `*text*`
  * **bold** `**text**`
  * sub~script~ `~text~`
  * super^script^ `^text^`
  * ~~strikethrough~~ `~~text~~`
  * "quotations"
  
## Spacing, manual breaks, lines
* line spacing effects
* two extra spaces at the end of a line for a manual break
* lists
  * indented
  * numbered

## Links
* links [linking text](website address)
* footnotes^2^
* images

## Tables


## subscripts in LaTex
$$H_0 = Z_{a + b}$$


## superscripts in LaTex 

$$S = cA^z$$

$$S=cA^z_1 + Z_{2 + x}$$
special characters proceeded with backslash

$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

$$z = \sum_{i=1}^X{k}$$

what if you need to include a backslash in the equation? There's a code for that

$$\backslash \alpha \le b \backslash$$

## Rendering in plain text

$$P(Occurrence of Species A) = Z$$
$$P(\mbox{Occurrence of Species A}) = Z$$

easy to embed LaTex in line $\pi + 3$ like so
we can also use markdown fencing in line with a single backtick `pi + 3` like this. We can even execute r code inside of markdown text `r pi + 3` << " r pi + 3 "

## R code in chunks 

```{r eval=TRUE, echo=TRUE}
# use comments extensively in all of your coding!
Pred <- seq(1,10) # make a vector of integers from 1 to 10
Res <- runif(10) # generate 10 draws from a random uniform (0,1) distribution

# print the random numbers
print(Res)

# plot the graph
plot(x=Pred,y=Res,type="b")

```

